---
title: "Data packages in R"
subtitle: "Statistical Programming"
author: 
  - "Alexa Fredston"
date: 'January 2022'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
seal: false
---
<style>
.title-slide {
vertical-align: bottom !important; 
text-align: left !important;
}
</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
library(tweetrmd) 
style_duo_accent(
  primary_color = "#EBB59F",
  secondary_color = "#0B1B32",
  text_color = "#0B1B32",
  inverse_header_color = "#0B1B32",
  title_slide_text_color="#0B1B32",
  link_color = "#EBB59F",
  header_font_google = google_font("Cardo"),
  text_font_google = google_font("Work Sans"),
  code_font_google = google_font("Fira Mono")
)
```

---

# What is a package? 

---

# What is a data package? 

---

# How data packages work


---

# Advantages of data packages

Strengthen reproducibility by making the data import process part of your code

Avoid errors from importing data incorrectly or using the wrong source data

Facilitate collaborators accessing and working with source data

---

# Simple data packages

### Useful for instruction and reproducible examples 

* [gapminder](https://github.com/jennybc/gapminder)
* [palmerpenguins](https://github.com/allisonhorst/palmerpenguins/)

---

# Environmental data packages

Most of the ones I use regularly are from [ROpenSci]()

### Some favorites

* [rnaturalearth](https://github.com/ropensci/rnaturalearth)
* [rgbif]
* [rerddap](https://github.com/ropensci/rerddap)
* [FishData](https://github.com/James-Thorson/FishData)

---

# rerddap

```{r, message=FALSE, echo=TRUE}
library(rerddap)
library(raster)
grid_out <- griddap("ncdcOisst2Agg_LonPM180", time=c("2020-03-10","2020-03-17"), latitude = c(39, 41), longitude = c(-74.5, -73.5), fields="sst")
grid_brick <- brick(grid_out$summary$filename)
plot(grid_brick)
```

---

# rnaturalearth

```{r, message=FALSE, echo=TRUE}
library(rnaturalearth)
plot(ne_coastline())
```

---

# rgbif

```{r, message=FALSE, echo=TRUE}
library(rgbif)
name <- name_suggest(q="Balaenoptera musculus")
name
key <- name$data$key[1]
occ_count(taxonKey=key, georeferenced=TRUE)
```

---

# Where to find them

[ROpenSci list](https://ropensci.org/packages/data-access/)

## Not all open datasets are in packages!

* [Awesome Public Datasets](https://github.com/awesomedata/awesome-public-datasets)
* [Ecological Data Wiki](https://ecologicaldata.org/)
* [OpenTraits](http://opentraits.org/)
* [Continuous Plankton Recorder](https://www.cprsurvey.org/data/our-data/)
* [Envirem](http://envirem.github.io/)
* [BioTime](https://biotime.st-andrews.ac.uk/)

---

# Caveats

Beware updates / versions (e.g., gapminder)

Can be buggy; check the GitHub Issues

Be mindful of data provenance 

Cite them, please!  `citation(package = 'rerddap')`

---

# Take-home messages

Data packages are a great way to make analysis easier and more reproducible 

Take a step back and check major R resources before spending a lot of time searching for or customizing anything


